---
title: "The summer break effects on the CoViD-19 pandemic in Germany 2020-2022"
author: 
  name: Jan Lorenz
  email: jlorenz@constructor.university
  orcid: 0000-0002-5547-7848
  affiliations:
    name: Constructor University
date: "Jan 10, 2023"
format: 
  revealjs:
    code-fold: true
    embed-resources: true
---

# Do summer breaks *increase* or *decrease* the pandemic?

More technical: Do the reduce or increase the reproductive number?


## Both directions possible

**Decreasing**  

☀️ Summer break ➡️ kids are not in school ➡️ parents take holidays ➡️ cuts off contagious channels

. . . 

**Increasing**  
☀️ Summer break ➡️ families start to travel ➡️ new channels of contagion meeting strangers, relatives, and friends

. . . 

**How can we study which effect is stronger?**

## Data

```{r}
#| label: packages-and-data-import
#| output: false
# These are the packages used in the report
library(tidyverse)
library(lubridate) # for date wrangling
library(tidymodels) # for hypothesis test
library(sf) # For map in appendix
library(patchwork) 

# Data
# raw data sources, download or documentation, data processing, and saving are done 
# by the script download_data.R

# Load the data produced by down_load.R
load("data/RKI_vacation_shp.RData")
# This loads
# RKIstate  : 16854 x 3 (stateCode, Refdatum, Fall)
#             Has new cases (Fall) per federal state (stateCode) per day (Refdatum)
# summervac : 48 x 5 
#             Summer school vacation per federal state (BL=shortcode, Land=name of state, 
#             with start and end date and year
# DE_shp    : Shape files for maps of federal states 
```

Detailed data for regional units in Germany is available through the Robert-Koch-Institut (RKI) at
<https://github.com/robert-koch-institut/SARS-CoV-2-Infektionen_in_Deutschland>. For this analysis the data is aggregated to new cases per day per federal state. For the following analyses we use seven-day averages to smooth out the typical weekly seasonality of CoViD-19 reporting. That means for each day we look at the average of the new cases of that day and the six days before that day. 


## German Trend

```{r}
smooth_cases <- function(x) (x + lag(x,1) + lag(x,2) + lag(x,3) + lag(x,4) + lag(x,5) + lag(x,6))/7
RKI_DE <- RKIstate |> group_by(Refdatum) |> 
  summarise(Fall_DE = sum(Fall, na.rm = TRUE)) |> 
  mutate(Fall_DE_smooth = smooth_cases(Fall_DE),
         year = year(Refdatum))
RKI_DE |> filter(month(Refdatum) %in% 6:9) |> 
  ggplot(aes(Refdatum,Fall_DE_smooth)) + geom_line() + facet_wrap(~year, scales = "free", nrow = 3) +
  theme_minimal() + expand_limits(y = 0) + labs(x="",y="")
```

General trend of new cases (7 day average) in summer months. **Note the different values on the y-axis scales!**

## Summer breaks differ!

```{r}
#| message: false

# Compute variation of start and end dates and overlap of summer breaks
variation <- summervac |> group_by(year) |> 
  summarize(start_variation = max(start) -min(start),
            end_variation = max(end) -min(end),
            overlap = min(end)-max(start)+1)
# Plot of durations
summervac |> pivot_longer(c(start,end)) |> 
  ggplot(aes(x = value, y = fct_rev(Land))) + 
  geom_point() + geom_line() + 
  facet_wrap(~year, scales = "free_x") + 
  scale_x_date(date_minor_breaks = "1 week") +
  labs(x = "", y = "") +
  theme_minimal(base_size = 16)
```

. . . 

**Strategy of analysis:** Judge the effect of a summer break in comparision to the German-wide trend.


## Idea 

```{r}
# Extend the dataset with vacation data
RKIstate_ext <- RKIstate |> 
  mutate(Fall_smooth = smooth_cases(Fall)) |> 
  left_join(RKI_DE, by = "Refdatum") |> 
  left_join(summervac, by = c("year", "stateCode" = "BL"))

# Make a dataset with the measure of vacation_effect per year-state for various delays 
# (the default one is five days delay, the others are computed for robustness checks)
# vacation_effect = factor from normalized German-wide trend to the state's trend 
#                   at the end of the assessment period (end + delay)
# The German-wide trend is normalized at the start of the assessment period (start + delay)
# Computation is done for the respective summer break vacation periods with several delay 
# "delays" from -42 up to 47 days. That way we have effects in a similar period directly before the summer break
# and directly after the summer break, as well as all values in between, 
# including 0 days and 5 days (which use to display the main result)

# A function to compute the vacation effects for a particular delay
make_delayed_vacation_effect <- function(delay) RKIstate_ext |> 
 filter(Refdatum==(start + delay)) |> 
 mutate(trend_factor = 1/Fall_DE_smooth*Fall_smooth)|> 
 select(stateCode,year,trend_factor) |> 
 right_join(RKIstate_ext |> filter(Refdatum==start + delay | Refdatum == end + delay), 
            by = c("stateCode", "year")) |> 
 mutate(Trend = Fall_DE_smooth * trend_factor) |> 
 filter(Refdatum == end + delay) |> 
 mutate(delay = delay)

# Create the data set of the vacation effect for several delays
RKI_mdel <- seq(-42,47,1) |> map(make_delayed_vacation_effect) |> list_rbind() |> 
 mutate(vacation_effect = Fall_smooth/Trend, 
        log_vacation_effect = log(vacation_effect), 
        dir_color = if_else(vacation_effect < 1, "down", "up"))
```

```{r}
#| fig-height: 2.4

# colors for arrows in figure
colors <- c("up" = "red", "down" = "darkgreen")

# Make dataset with a normalized trend for each state with delay 5
RKI <- RKIstate_ext |>
  filter(Refdatum==start + 5) |>
  mutate(trend_factor = 1/Fall_DE_smooth*Fall_smooth) |>
  select(stateCode,year,trend_factor) |>
  right_join(RKIstate_ext, by = c("stateCode", "year")) |>
  mutate(Trend = Fall_DE_smooth * trend_factor)
RKI |> group_by(year, stateCode) |> filter(stateCode == "BW", month(Refdatum) %in% 6:9) |> 
  ggplot(aes(x = Refdatum)) +
  geom_rect(data = summervac |> filter(BL == "BW") |> mutate(Refdatum = start), 
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), fill = "gray", alpha = 0.5) +
  geom_rect(data = summervac |> filter(BL == "BW") |> mutate(Refdatum = start), 
            aes(xmin = end, xmax = end + 5, ymin = -Inf, ymax = Inf), fill = "gray", alpha = 0.25) +
  geom_point(data = RKI |> filter(stateCode == "BW", Refdatum == start + 5),
             aes(x = Refdatum, y = Fall_smooth), color = "blue", alpha = 0.5, size = 3) +
  geom_point(data = RKI |> filter(stateCode == "BW", Refdatum == start + 5),
             aes(x = Refdatum, y = Fall_smooth), size = 1.5) +
  geom_line(aes(y = Fall_smooth), color = "black") +
  geom_line(aes(y = Trend),color = "blue", alpha = 0.5) + 
  geom_segment(data = RKI_mdel |> filter(stateCode == "BW", delay == 5), 
               aes(yend = Fall_smooth, y = Trend, xend = Refdatum, x = Refdatum, color = dir_color),  
               arrow = arrow(length = unit(0.25,"cm"), type = "closed", angle = 20) ) +
  scale_color_manual(values = colors) +
  facet_wrap(~year, scales = "free") + 
  expand_limits(y = 0) + 
  labs(x="",y="", title = "Summer breaks in Baden-Württemberg") + 
  guides(color = "none") + 
  theme_minimal()
```


- Five days delay (incubation period)
- Re-scale the German-wide trend 
- Measure the summerbreak effect 5 days after the end of the summer break

## Vacation more often increase cases 

```{r}
RKI_mdel |> filter(delay == 5) |> 
  mutate(eff_ch = vacation_effect-1,
         lab_x = pmax(0,eff_ch),
         eff_lab = paste0(if_else(vacation_effect>1,"+",""),round(100*(vacation_effect-1), digit=1),"%")) |> 
  ggplot(aes(eff_ch, fct_rev(Land), label = eff_lab, fill = dir_color)) + 
  geom_col() +
  geom_text(aes(x=lab_x), hjust=0, nudge_x = 0.2) +
  facet_wrap(~year) +
  xlim(c(-0.5,7.5)) +
  scale_fill_manual(values = colors) + 
  labs(x="", y="", title = "Summer break effect") +
  guides(fill="none") +
  theme_minimal(base_size = 16) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_blank())
```

- In most federal states case number increased more than the German-wide trend
- The relationsships vanishes in 2022 (transition to endemic phase)

## Statistical significance 

```{r}
tt20 <- RKI_mdel |> filter(year == 2020, delay == 5) |> 
 t_test(response = log_vacation_effect, alternative = "greater")
tt21 <- RKI_mdel |> filter(year == 2021, delay == 5) |> 
 t_test(response = log_vacation_effect, alternative = "greater")
tt22 <- RKI_mdel |> filter(year == 2022, delay == 5) |> 
 t_test(response = log_vacation_effect, alternative = "greater")

# function to transform log back to percentage change
tr <- function(x) format(100*(exp(x)-1), digits = 3)
```

Year          | Mean effect            | Lower bound 0.95-CI        | p-value
--------------|-----------------------:|---------------------------:|-------------------------------------:
2020          | `r tr(tt20$estimate)`% | `r tr(tt20$lower_ci)`%     | `r format(tt20$p_value, digits = 3)`
2021          | `r tr(tt21$estimate)`% | `r tr(tt21$lower_ci)`%     | `r format(tt21$p_value, digits = 3)`
2022          | `r tr(tt22$estimate)`% | `r tr(tt22$lower_ci)`%     | `r format(tt22$p_value, digits = 3)`

t-test of value being greater than zero.



# Appendix: Additional analyses {.appendix}

## All federal states in 2020

```{r}
#| fig-height: 7.5
#| fig-width: 10
plot_year <- function(y,d) {
 RKI <- RKIstate_ext |>
  filter(Refdatum==start + d) |>
  mutate(trend_factor = 1/Fall_DE_smooth*Fall_smooth) |>
  select(stateCode,year,trend_factor) |>
  right_join(RKIstate_ext, by = c("stateCode", "year")) |>
  mutate(Trend = Fall_DE_smooth * trend_factor)
 RKI |> group_by(year, stateCode) |> filter(year == y, month(Refdatum) %in% 6:9) |> 
  ggplot(aes(x = Refdatum)) +
  geom_rect(data = summervac |> filter(year == y) |> mutate(stateCode = BL, Refdatum = start), 
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), fill = "gray", alpha = 0.5) +
  geom_rect(data = summervac |> filter(year == y) |> mutate(stateCode = BL, Refdatum = start), 
            aes(xmin = end, xmax = end + d, ymin = -Inf, ymax = Inf), fill = "gray", alpha = 0.25) +
  geom_point(data = RKI |> filter(year == y, Refdatum == start + d),
             aes(x = Refdatum, y = Fall_smooth), color = "blue", alpha = 0.5, size = 3) +
  geom_point(data = RKI |> filter(year == y, Refdatum == start + d),
             aes(x = Refdatum, y = Fall_smooth), size = 1.5) +
  geom_line(aes(y = Fall_smooth), color = "black") +
  geom_line(aes(y = Trend),color = "blue", alpha = 0.5) + 
  geom_segment(data = RKI_mdel |> filter(year == y, delay == d), 
               aes(yend = Fall_smooth, y = Trend, xend = Refdatum, x = Refdatum, color = dir_color),  
               arrow = arrow(length = unit(0.25,"cm"), type = "closed", angle = 20) ) +
  scale_color_manual(values = colors) +
  facet_wrap(~Land, scales = "free", nrow = 4) + 
  expand_limits(y = 0) + labs(x="",y="", title=paste("Summer breaks",y)) + guides(color = "none") + 
  theme_minimal()
}
plot_year(2020, 5)
```

## All federal states in 2021

```{r}
#| fig-height: 7.5
#| fig-width: 10
plot_year(2021, 5)
```

## All federal states in 2022

```{r}
#| fig-height: 7.5
#| fig-width: 10
plot_year(2022, 5)
```



## No geographic pattern {.appendix}

```{r}
DE_shp |> right_join(RKI_mdel |> filter(delay == 5), by = c("NAME_1" = "Land")) |> 
  ggplot(aes(fill=100*(vacation_effect-1))) + geom_sf() +
  facet_wrap(~year) +
  scale_fill_gradient2(
    low = "darkgreen",
    mid = "white",
    high = "red",
    midpoint = 0,
    name = "Change (%)",
    limits = c(-50,200),
    na.value = "red"
  ) +
  theme_void() +
 labs(caption = "Color axis capped at +200%. 3 states above in 2020 in dark red. ")
```

## Simulation null distributions and plotting the observed average effect {.appendix}

```{r}
#| fig-height: 3
plot_null_distribution <- function(y) {
 set.seed(1)
 point_estimate <- RKI_mdel |> filter(year == y, delay == 5) |> 
  specify(response = log_vacation_effect) |> 
  calculate("mean")
 RKI_mdel |> filter(year == y, delay == 5) |> 
  specify(response = log_vacation_effect) |> 
  hypothesize(null = "point", mu = 0) |> 
  generate(reps = 5000, type = "bootstrap") |> 
  calculate(stat = "mean") |> 
  visualize() + 
  shade_p_value(obs_stat = point_estimate, direction = "greater") + 
  theme_minimal() + labs(title = y)
}
plot_null_distribution(2020) + plot_null_distribution(2021) + 
plot_null_distribution(2022) + plot_annotation(title = "Simulation-based Null Distributions")
```


::: aside 
Bootstrapped null distributions for the samples of the log-transformed effect for 2020, 2021, and 2022. For each sample there 16 values (one per federal state). The null hypothesis is that the mean of these 16 values is zero. The red line shows the point estimate of the observed means.
:::